Tree height and hydraulic traits shape growth responses across droughts in a temperate broadleaf forest

### 2020-02-03 manuscript/climate_sensitivity_draft.Rmd
editing methods & results

We analyzed tree-ring data (cambial growth increment) from 571 trees representing the twelve species contributing most to woody aboveground net primary productivity ($ANPP_{stem}$), which together comprised 97% of study plot $ANPP_{stem}$ between 2008 and 2013 [@helcoski_growing_2019] (Figure S1). Cores were collected within the ForestGEO plot at breast height (1.3m) in 2010-2011 or 2016-2017 using a 5mm increment borer. In 2010-2011, cores were collected from randomly selected live trees of species with at least 30 individuals $\ge$ 10cm DBH [@bourg_initial_2013]. In 2016-2017, cores were collected from all trees found dead in the annual mortality census [@gonzalezakre_patterns_2016]. Cores were sanded, measured, and cross-dated using standard procedures, as detailed in [@helcoski_growing_2019]. The resulting chronologies have been published in association with @helcoski_growing_2019: (ITRDB; GitHub/Zenodo).  **Ryan, have you heard back from ITRDB?  We should also cite GitHub/Zenodo here. **

@@

@@

*Tree size and drought resistance*

Overall, our analysis confirmed the tendency for larger-diameter trees to show greater reductions in growth during drought [@bennett_larger_2015] (*H1.0*), although there was no significant effect during 1977 or 1999 individually (Tables 1, 4). The same held true for $ln[H]$ as a univariate predictor (*H1.1*; Tables 1, 4). When combined with other predictor variariables in our multivariate models, the top models usually included an effect of $ln[H]$, and its coefficient was consistently negative, as predicted (Tables 1, 5). We note that a non-signficant positive correlation between $ln[H]$ and $Rt$ for 1999 became negative in the context of the multivariate models, again supporting *H1.1* (Table 1).

Crown position was generally correlated with $H$, but with substantial variation (Fig. 2d). Crown position was a much poorer predictor of $Rt$ than was $H$ (Table 4), lending little overall support to *H1.2* (Table 1). When considered alone, $CP$ had a significant response only in the 1966 drought, during which trees with dominant $CP$ had the lowest $Rt$. When considered in conjunction with $H$, $CP$ resulted as a significant predictor only for the 1999 drought, during which suppressed and then intermediate trees had the lowest $Rt$. Crown position was included in almost half of the top models, with mixed results as to how $Rt$ varied with $CP$ (Table 5). Most commonly in these multivariate models, as in the univariate models (Table 4), the resistance of dominant trees was less than that of co-dominant trees but higher that of suppressed trees. Although $CP$ was a useful predictor of $Rt$, it overall had a weak effect relative to that of $H$.

@@

Resistance was negatively correlated with $ln[TWI]$ (Tables 4-5), negating the hypothesis that trees in moist microsites would suffer less during drought. Nevertheless, we tested for a negative $ln[H] *ln[TWI]$ interaction (*H1.3*), which could indicate that smaller trees (with smaller rooting volume) are more susceptible to drought in drier microenvironments that have a deeper water table. *H1.3* was rejected; the $ln[H] *ln[TWI]$ interaction was never signficant and had a consistently positive coefficient (Table 4).

*Species' traits, height, and drought resistance*

@@

The results reject Hypothesis *H2.2*, finding no evidence that taller trees had traits associated with greater drought resistance. In part because of the large sample size (n=7549 trees--all individuals of our 12 focal species $\ge$ 10 cm in 25.6 ha), there were significant (p<0.0001) correlations of $H$ with all species' traits (see Table S4).**see comment 35** $WD$ and $LMA$ were the only traits where the correlation matched the predicted direction (*i.e.*, evidence that more drought-resistant traits are associated to taller trees), but ultimately neither of these traits were useful predictors of $Rt$. Furthermore, although correlations were statistically significant, trait variability within each height class overwhelmed any vertical trends (Fig. 2e-f).

Our findings support Hypothesis *H2.3* that the observed tendency for larger trees to have greater growth reductions during drought (lower $Rt$) is driven by height itself, as opposed to more drought-sensitive traits in larger trees (Tables 1,5). As discussed above, there was little meaningful variation in traits with height at the community level. When $ln[H]$ and hydraulic traits were considered together in multivariate models, the effect of $ln[H]$ on $Rt$ was consistently negative (Table 5). Simultaneously in the 1999 univariate model, this co-consideration resulted in a non-signficant positive $ln[H]$-$Rt$ correlation (Table 4).

*Responses across droughts*

Our results reject the hypothesis that overall community responses varied across droughts (*H3.1*). Within the context of mixed effects models, there were no significant differences in $Rt$ across drought years (Table 4). **see comment 36**

@@

Further, our findings support the hypothesis (*H3.3*) that the strength of predictor variables was different across the droughts (Tables 1,4,5). For instance, $ln[H]$ and $PLA_{dry}$ had much stronger negative effects in 1966 than in the other two years, $ln[TWI]$ had the strongest negative effect in 1977, and $XP$ (higher $Rt$ among ring-porous trees) was strongest in 1999 (Tables 4,5). **see comment 37**

>>>>

We analyzed tree-ring data (cambial growth increment) from 571 trees representing the twelve species contributing most to woody aboveground net primary productivity ($ANPP_{stem}$), which together comprised 97% of study plot $ANPP_{stem}$ between 2008 and 2013 [@helcoski_growing_2019] (Figure S1). Cores were collected within the ForestGEO plot at breast height (1.3m) in 2010-2011 or 2016-2017 using a 5mm increment borer. In 2010-2011, cores were collected from randomly selected live trees of species with at least 30 individuals $\ge$ 10cm DBH [@bourg_initial_2013]. In 2016-2017, cores were collected from all trees found dead in the annual mortality census [@gonzalezakre_patterns_2016]. Cores were sanded, measured, and cross-dated using standard procedures, as detailed in [@helcoski_growing_2019]. The resulting chronologies were published in association with @helcoski_growing_2019 (DOI: 10.5281/ zenodo.2649302).

@@

All data, code, and results are available through the SCBI-ForestGEO organization on GitHub (https://github.com/SCBI- ForestGEO: SCBI-ForestGEO-Data and McGregor_climate-sensitivity-variation repositories), with static versions corresponding to data and analyses presented here archived in Zenodo (DOIs: 10.5281/zenodo.3604993 and [TBD], respectively).

@@

*Tree size, microenvironment, and drought resistance*

Overall, our analysis confirmed the tendency for larger-diameter trees to show greater reductions in growth during drought [@bennett_larger_2015], although there was no significant effect during 1977 or 1999 individually (Tables 1, 4). The same held true for $ln[H]$ as a univariate predictor (*H1a*; Tables 1, 4). When combined with other predictor variariables in our multivariate models, the top models usually included an effect of $ln[H]$, and its coefficient was consistently negative, as predicted (Tables 1, 5). We note that a non-signficant positive correlation between $ln[H]$ and $Rt$ for 1999 became negative in the context of the multivariate models, again supporting *H1a* (Table 1).

Crown position was generally correlated with $H$, but with substantial variation (Fig. 2d). Crown position was a much poorer predictor of $Rt$ than was $H$ (Table 4), lending little overall support to *H1b* (Table 1). When considered alone, $CP$ had a significant response only in the 1966 drought, during which trees with dominant $CP$ had the lowest $Rt$. When considered in conjunction with $H$, $CP$ resulted as a significant predictor only for the 1999 drought, during which suppressed and then intermediate trees had the lowest $Rt$. Crown position was included in **almost half of the top models, with mixed results as to how $Rt$ varied with $CP$ (Table 5). Most commonly in these multivariate models, as in the univariate models (Table 4), the resistance of dominant trees was less than that of co-dominant trees but higher that of suppressed trees**. Although $CP$ was at times a useful predictor of $Rt$, it overall had a weak effect relative to that of $H$.

@@

Resistance was negatively correlated with $ln[TWI]$ (Tables 4-5), negating the hypothesis that trees in moist microsites would suffer less during drought. Nevertheless, we tested for a negative $ln[H] *ln[TWI]$ interaction (*H1c*), which could indicate that smaller trees (with smaller rooting volume) are more susceptible to drought in drier microenvironments that have a deeper water table. *H1c* was rejected; the $ln[H] *ln[TWI]$ interaction was never signficant and had a consistently positive coefficient (Table 4).

*Species' traits and drought resistance*

@@

*Combined effects of tree size and traits*

*Our findings indicate that the greater growth reductions in large trees during drought (lower $Rt$) is driven by height itself, as opposed to more drought-sensitive traits in larger trees (Table 5). Tthere was little meaningful variation in traits with height at the community level (**Fig. S?**). When $ln[H]$ and hydraulic traits were considered together in multivariate models, the effect of $ln[H]$ on $Rt$ was consistently negative (Table 5). *

*Responses across droughts* {**remove this section, but first see if there's anything that should be preserved.**}

@@

 For instance, $ln[H]$ and $PLA_{dry}$ had much stronger negative effects in 1966 than in the other two years, $ln[TWI]$ had the strongest negative effect in 1977, and $XP$ (higher $Rt$ among ring-porous trees) was strongest in 1999 (Tables 4,5). **see comment 37**
